---
title: "project visualisation"
author: "Yara Roelen"
date: "2025-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rsconnect)
library(ggplot2)
library(ggmap)
library(ggspatial)
library(sf)
library(raster)
library(geodata)
library(leaflet)
library(dplyr)
library(RColorBrewer)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(corrplot)
library(GGally)
```
# read the data from the different stations
```{r}
data_stations = read.csv("C:/users/yarar/downloads/unzipped/stations.csv")
head(data_stations)
data_stations = na.omit(data_stations)
```
# read the data of all the different years and paste them all under each other
```{r}

years <- 2001:2018


base_path <- "C:/users/yarar/downloads/unzipped/csvs_per_year/csvs_per_year/"

# read data and paste under each other
all_data <- lapply(years, function(year) {
  file_path <- paste0(base_path, "madrid_", year, ".csv")
  df <- read.csv(file_path)
  print(df)
  df$year <- year  # add year as a column
  return(df)
}) %>% bind_rows()
# the name of station is changed to id
colnames(all_data)[colnames(all_data) == "station"] <- "id"
# the data of the years is mergend with the data of the stations
merged_all_data <- merge(all_data, data_stations, by = "id")

```
# summarize the data (take averages for each pollutant)
```{r}
averages_all_data = all_data %>% group_by(year) %>% summarize(mean_nitrogen = mean(NO_2, na.rm = TRUE))
```


# Missing values
```{r}
# number of missing values
colSums(is.na(all_data))
# total numbber of rows
colSums(is.na(all_data)) + colSums(!is.na(all_data))

# missing variables per year and per variable
missing_stats_per_year_var <- all_data %>%
  group_by(year) %>%
  summarise(across(everything(), list(
    missing = ~ sum(is.na(.)),
    total = ~ n()
  ))) %>%
  pivot_longer(
    cols = -year,
    names_to = c("variable", "stat"),
    names_pattern = "^(.*)_(missing|total)$"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>%
  mutate(percentage_missing = (missing / total) * 100)

```
# Missing values per station in 2018
```{r}
library(dplyr)
library(tidyr)

vars <- c("BEN", "CO", "EBE", "NO", "NOx", "NO_2", "O_3", "PM10", "PM25", "SO_2", "TCH", "TOL", "PXY", "MXY", "NMHC", "OXY", "CH4")

missing_summary_2018 <- merged_all_data %>%
  filter(year == 2018) %>%
  # alleen de kolommen year, name + de pollutants behouden
  select(year, name, all_of(vars)) %>%
  pivot_longer(
    cols      = all_of(vars),
    names_to  = "variable",
    values_to = "value"
  ) %>%
  group_by(name, variable) %>%
  summarise(
    n_missing    = sum(is.na(value)),
    total        = n(),
    perc_missing = 100 * n_missing / total,
    .groups      = "drop"
  )
```
# variables with more then 99% of missing in a station
```{r}
vars_99plus_missing <- missing_summary_2018 %>%
  filter(perc_missing > 99) %>%
  distinct(variable)
```

# summary measures NO2 2018
```{r}
# summary statistics of NO2 for each station
no2_stats_2018 <- merged_all_data %>%
  filter(year == 2018) %>%
  group_by(name) %>%
  summarise(
    mean_NO2 = mean(NO_2, na.rm = TRUE),
    median_NO2 = median(NO_2, na.rm = TRUE),
    sd_NO2 = sd(NO_2, na.rm = TRUE),
    .groups = "drop"
  )

# boxplot of NO2 for each station
merged_all_data %>%
  filter(year == 2018, !is.na(NO_2)) %>%
  ggplot(aes(x = reorder(name, NO_2, FUN = median, na.rm = TRUE), y = NO_2)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.size = 1) +
  labs(
    title = "Boxplot of NO_2 per station in 2018",
    x = "Station",
    y = expression(NO[2] ~ (µg/m^3))
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


# General summary measures
```{r}
library(dplyr)
library(tidyr)

# summarize data
summary_overall <- all_data %>%   
  select(all_of(vars)) %>%         
  summarise(
    across(
      everything(),
      list(
        mean = ~mean(., na.rm = TRUE),
        median = ~median(., na.rm = TRUE),
        sd = ~sd(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )


# pivot to longer
summary_long <- summary_overall %>%
  pivot_longer(
    everything(),
    names_to = c("variable", "stat"),
    names_sep = "_(?=[^_]+$)",  # split on last underscore
    values_to = "value"
  )

# check on duplicates
duplicates <- summary_long %>%
  summarise(count = n(), .by = c(variable, stat)) %>%
  filter(count > 1)

# back to wider but without the duplicates
summary_tidy <- summary_long %>%
  pivot_wider(
    names_from = stat,
    values_from = value,
    values_fn = mean  
  )



```
# summary statistics per year
```{r fig.width=14, fig.height=25}
library(dplyr)
library(tidyr)

summary_per_year <- all_data %>%
  group_by(year) %>%  
  summarise(
    across(
      all_of(vars),  
      list(
        mean = ~mean(., na.rm = TRUE),
        median = ~median(., na.rm = TRUE),
        sd = ~sd(., na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  ungroup()


# include year in the output
summary_long_per_year <- summary_per_year %>%
  pivot_longer(
    cols = -year,
    names_to = c("variable", "stat"),
    names_sep = "_(?=[^_]+$)",
    values_to = "value"
  )


```
# scatter plots
```{r}
library(tidyverse)

# pollutants
vars <- c("BEN", "CO", "EBE", "NO", "NOx", "NO_2", "O_3", "PM10", "PM25", "SO_2", "TCH", "TOL", "PXY", "MXY", "NMHC", "OXY", "CH4")

long_data_year <- all_data %>%
  select(year, all_of(vars)) %>%
  pivot_longer(cols = -year, names_to = "variable", values_to = "value")

# Plot each variable against year
ggplot(long_data_year, aes(x = year, y = value)) +
  stat_summary(fun = mean, geom = "line") +       # line with average per year
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2) +
  facet_wrap(~variable, scales = "free_y", ncol = 4) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) + theme(axis.text.y = element_text(size = 4)) +
  labs(title = "Trends of pollutants over years",
       y = "Mean concentration",
       x = "Year")

```


```{r fig.width=14, fig.height=25}

long_data_station <- merged_all_data %>%
  select(name, all_of(vars)) %>%
  pivot_longer(cols = -name, names_to = "variable", values_to = "value")

# Plotting: boxplot per station per variable
ggplot(long_data_station, aes(x = factor(name), y = value)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  theme_minimal() +
  labs(title = "Pollutant concentrations per station",
       y = "Concentration",
       x = "Station") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text.y = element_text(size = 10)) + theme(axis.text.x = element_text(size = 10))

```


# trend over time for each pollutant
```{r}
all_data %>%
  group_by(year) %>%
  summarise(across(c(BEN, CO, EBE, NO, NOx, NO_2, O_3, PM10, PM25, SO_2, TCH, TOL, PXY, MXY, NMHC, OXY, CH4), mean, na.rm = TRUE)) %>%
  pivot_longer(-year, names_to = "pollutant", values_to = "mean_value") %>%
  ggplot(aes(x = year, y = mean_value, color = pollutant)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Average pollutant levels per year") + 
  ylab("Mean pollutant level")


```
# correlations
```{r}
# Select relevant numeric columns
num_vars <- all_data %>%
  dplyr::select(BEN, CO, EBE, NO, NOx, NO_2, O_3, PM10, PM25, SO_2, TCH, TOL, PXY, MXY, NMHC, OXY, CH4)
num_vars <- num_vars %>% rename_with(make.names)
num_vars <- as.data.frame(num_vars)


# Correlation matrix
corrplot(cor(num_vars, use = "pairwise.complete.obs"), method = "color", type = "upper")

```


